### 2019-07-06 Memoria/Chapters/ConstrainedClustering.tex
Cambios en memoria y en paper DILS

Once the basics concepts about clustering have been introduced, we can focus im defining the concepts related to constrained clustering, as well as to give examples of its application and highlight the benefits and problems presented by its use. To this end we take as our main reference the survey presented in \cite{davidson2007survey}. We will start by describing the motivation for constrained clustering in Section \ref{sec:CCMotiv}. We formalize the concept of constraints and discuss their use in Sections \ref{sec:ConstDef} and \ref{sec:ConstUse} respectively. Afterwards a list of constrained clustering application is presented in Section \ref{sec:CCApplications}. Benefits and problems, as well as solutions for these problems, are discussed in Sections \ref{sec:ConstraintsBenefits} and \ref{sec:ConstraintsProblems} respectively.

@@

As we discussed in Chapter \ref{ch:IntroClustering}, unsupervised clustering methods are useful for structuring data referring to a particular area. An example of this can be found in text classification; In \cite{cohn2003semi} a problem proposed by Yahoo! is addressed. The problem consist of, given a large number of text documents, grouping them according to a taxonomy in which documents with similar topics are conceptually nearby. To solve this problem, unsupervised clustering methods are useful, as the information on the problem initially available is limited. However, in \cite{wagstaff2001constrained} is shown that applying unsupervised clustering to certain problems, such as grouping \acf{GPS} data in such a way that clusters define the lanes of a road, does not produce significant results, as the clusters obtained are far from the elongated shape that would be expected as a result. To tackle the problem, they introduced a new element in clustering, the constraints, which made it possible to include domain knowledge that would guide the clustering methods to obtain the expected results. It was enough to indicate that the lanes of the road on which the vehicles circulate are four meters wide, and therefore any vehicle that is at a distance of more than 4 meters from another, perpendicular to the motion direction, must be located in a different cluster.

We are now in a new scenario: it is possible to incorporate additional information to the clustering process, in addition to that contained in the dataset itself, to guide it in the formation of the output partition and obtain more accurate results. This new approach is known as constrained clustering, a \acf{SSL} paradigm, unlike traditional clustering methods which fall within the area of unsupervised learning.

@@

The new type of information that we incorporate to clustering is given in the form of instance-level constraints, that is, to specify if two instances ($x_i$ and $x_j$) from the dataset ($X$) must be in the same cluster or, on the contrary, they must be in separate clusters.

@@

Although they may seem simple, constraints defined as above have some very interesting properties. \acs{ML} constraints are an example of an equivalence relationship, and therefore are symmetrical, reflexive and transitive. Observation \ref{ob:MLConstraintsRelation} formalizes this concept. On the other hand, and although \acs{CL} do not constitute an equivalence relation, they feature other properties, formalized in Observation \ref{ob:CLConstraintsRelation}.

@@

A clear example of constraint use can be found in clustering applications where there are distance measurement limitations, as in the case of \acs{GPS} data. Thus, if we want that instances forming two clusters to be separated by a distance greater or equal to $\delta$, it is enough to set \acs{ML} constraints between all instances separated by less than $\delta$ units. Similarly, if we want the diameter of the clusters to be at most $\epsilon$, we must set \acs{CL} constraint between instances separated by more than $\epsilon$ units. Figure \ref{fig:DistanceConstraints} shows a graphical representation of these two types of constraints.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.45]{gfx/ConstClust/RestriccionesDeltaEpsilon.png} 
	\caption[\textit{delta} and \textit{epsilon} constraints]{\textit{delta} and \textit{epsilon} constraints \cite{davidson2007survey}.}\label{fig:DistanceConstraints}
\end{figure}

@@

While supervised learning involves knowing the label associated with each instance, \acs{SSL} has only a subset of labeled instances. On the other hand, in many domains the available information refers to relations between instances, and not to the specific class to which they belong. Moreover, in interactive clustering systems, a user who is not expert in the problem domain will probably be able to provide information in the form of constraints such as \acs{ML} and \acs{CL}, rather than providing information on what particular class certain instances belong to \cite{cohn2003semi,davidson2007hierarchical}.

Typically, constraints are incorporated into clustering problems in two ways. They can be used to alter the instance clusters assignment rule of the method in hand, so that the solution satisfies as many constraints as possible. Alternatively, it is possible to train the distance function used by the method based on the constraints, either before or during the application of the method. In any case, the initialization phase can take the constraints into account, so that instances associated with constraints \acs{ML} will be placed in the same cluster, and those between which there is a constraint \acs{CL} will be placed in different clusters. Based on this distinction, we identify two ways of approaching the problem, those based on explicit constraints (\textit{constraints-based}), and those based on distances (\textit{distance-based}).

@@

	\item Initialize clusters based on constraints inferred from a set of labeled instances \cite{basu2002semi}.

@@

	\caption[Constraints and clustering over a dataset]{Constraints and clustering over a dataset \cite{davidson2007survey}.} \label{fig:ConstClustOverDataset}
\end{figure}

@@

In distance-based approaches, classic clustering methods that use a distance measurement are employed, so that the distance measurement is modified to incorporate the constraints. In this context, satisfying constraints means that instances related to constraints \acs{ML} are placed together in the space, and those related by \acs{CL} are separated.

Figure \ref{fig:ConstrainsAndMetricLearned} shows a possible clustering based on a metric learned from the constraints. It should be noted that in Figure \ref{fig:MetricLearned} the data space has been compressed on the vertical axis and widened on the horizontal axis to match the learned distance metric.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Constraints over a dataset.]
	{\includegraphics[width=.45\linewidth]{gfx/ConstClust/InputInstancesAndConst2}} 
	\hspace{0.5cm}
	\subfloat[Clustering based on a metric learned from constraints.]
	{\includegraphics[width=.45\linewidth]{gfx/ConstClust/MetricaAprendida}
	\label{fig:MetricLearned}}
	\caption[Clustering based on a metric learned from constraints.]{Clustering based on a metric learned based on constraints \cite{davidson2007survey}.} \label{fig:ConstrainsAndMetricLearned}
\end{figure}

@@

This section shows some application cases in which constrained clustering has turned out to be a more useful tool than unsupervised clustering. For each case we will analyze how the constraints were obtained and how they improve the results in the resulting clustering.

@@

Figure \ref{fig:CMUFacesDatabase} shows a sample from the CMU (Carnegie Mellon University) faces dataset, where the task is to group faces based on different criteria. In this case, the goal is to group faces according to their orientation.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Profile faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras1}} \quad
	\subfloat[Front faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras3}} \quad
	\subfloat[Faces up.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras2}} \quad
	\caption[CMU faces database.]{CMU faces database \cite{davidson2007survey}.}\label{fig:CMUFacesDatabase}
\end{figure}

The method used to obtain the constraints is one of the most popular in the literature: set the number of clusters of the resulting partition equal to the number of classes in the database, and generate the constraints from a subset of labeled instances; that is, if two instances have different labels, set a \acs{CL} constraint between them, otherwise one of \acs{ML} type. Thus, between the images shown in Figure \ref{fig:FacesDatabaseCL} \acs{CL} constraints are set, since, although they belong to the same person, they do not have the same orientation.

@@

Figure \ref{fig:AiboRobotClustSys} shows another set of image data on which constrained clustering techniques are applied. In this case, the task is to perform object recognition to incorporate the method into the navigation system of the Aibo robot \cite{davidson2005clustering}. Distance constraints such as $\delta$ and $\epsilon$ are used as described in Figure \ref{fig:DistanceConstraints}. In this way well differentiated clusters are obtained and therefore they are useful for the path-finding techniques performed by the robot during navigation.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Imagen original]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo1}} \quad
	\subfloat[Clustering sin restricciones]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo2}} \quad
	\subfloat[Clustering con restricciones]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo3}} \quad
	\caption[Clustering method used in Aibo robot navigation system.]{Clustering method used in Aibo robot navigation system \cite{davidson2007survey,davidson2005clustering}.}\label{fig:AiboRobotClustSys}
\end{figure}

@@

Video databases are one example where constraints can be generated directly from the data domain, especially if space-time data information is available \cite{yan2006discriminative}. In time sequenced data it is possible to set \acs{ML} constraints between groups of pixels of frames close in time. This is especially useful when the task is to perform object recognition based on clustering and segmentation. It is also possible to add \acs{CL} constraints to image segment in the same snapshot, as there is a low probability that they are associated with the same object after segmentation is done. In fact, in the video analysis domain problems there are a variety of constraint extraction methods \cite{yan2006discriminative}, Figure \ref{fig:ConstExtractionVideo} shows some examples.

@@

In Figure \ref{fig:ConstExtractionVideo}, image \ref{fig:ConstExtractionVideoA} corresponds to constraints extracted from tracking a person over a period of time, \ref{fig:ConstExtractionVideoB} corresponds to spatial constraints associating two objects located in the same frame, image \ref{fig:ConstExtractionVideoC} corresponds to restrictions obtained through facial recognition and \ref{fig:ConstExtractionVideoD} to those provided by the user.

With so many constraint extraction methods, one might ask: what happens if too many constraints are imposed? Does this make the problem over-constrained? In Section \ref{sec:ConstraintsProblems} we will address these questions.

@@

In gene clustering based on micro-array data, genes are represented by their expression profile in different experiments and grouped using different constrained clustering methods. Figure \ref{fig:GeneticApp} shows an example: these are \acs{ML} constraints set between genes based on co-occurrence data stored in the protein interaction database, which contains information about which genes (and their associated proteins) are present in the same cell processes \cite{xenarios2001dip}. This information can be used to improve results provided by clustering methods. \cite{segal2003discovering}.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/Genetica/Genes} 
	\caption[Gene clustering based on micro-array data.]{Gene clustering based on micro-array data \cite{davidson2007survey}.}\label{fig:GeneticApp}
\end{figure}

@@

In content classification tasks, the goal is to automatically split large amounts of documents into groups or clusters. In this case it is possible to extract constraints from multiple auxiliary resources. For example, if two documents are in the same directory, a \acs{ML} constraints could be set between them. This way it is possible to adjust the resulting clustering to suit a particular criterion, such as creating a hierarchy of documents similar to the way they are organized in the input directories structure.

@@

Constrained clustering is very useful in web search data processing. Here, the aim is to automatically group the results of an ambiguous query in the search engine into clusters of \acs{URL}s, which refer to the concept introduced as a query in different contexts. In this area it is possible to extract the constraints from searches previously performed by users, so that a \acs{ML} constraints is set between \acs{URL}s visited in the same user session. Clustering using these constraints can help bias the result of searches towards user's preferences.

@@

Constrained clustering is very useful in the processing of search data on web pages. Here, the aim is to automatically group the results of an ambiguous query in the search engine into clusters of \acs{URL}s, which refer to the concept introduced as a query in different contexts. In this area it is possible to extract constraints from searches previously performed by users, so that a \acs{ML} constraint is set between visited \acs{URL}s in the same user session. Clustering using these constraints can help biasing search results towards user preferences.

\subsection{GPS Data} \label{sec:GPSApp}

As mentioned at the beginning of Chapter \ref{ch:ConstrainedClustering}, constrained clustering can be applied to \acs{GPS} data to identify the lane in which each vehicle is running, as shown in Figure \ref{fig:GPSClustData}. Each instance is represented by the position it occupies on the road in two-dimensional Cartesian coordinates $(x,y)$, obtained from \acs{GPS} data. Figure \ref{fig:figure15} graphically shows this representation of the data (note that multiple instances can refer to the same vehicle at different times).

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/GPS/Coches} 
	\caption[GPS data use.]{\acs{GPS} data use. \cite{davidson2007survey} \cite{wagstaff2001constrained}}\label{fig:GPSClustData}
\end{figure}

In this domain, real clusters have an elongated shape on the horizontal axis and are aligned perpendicularly to the motion direction. To get the resulting clusters to have this shape, we can make use of the constraints. We will set a \acs{CL} constraint between those instances more than four meters apart perpendicularly to the motion direction (since the lanes have a maximum width of four meters), and \acs{ML} constraints between those instances that present continuity in the motion direction axis, since it is likely that the vehicles they represent are in the same lane. This clustering model has proven to be very useful in real-time navigation \cite{wagstaff2001constrained}, allowing the user to be notified when to change lanes, or when not to leave them.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.32]{gfx/ConstClust/GPS/Instancias} 
	\caption[Clustering of GPS data with no constraints.]{Clustering of \acs{GPS} data with no constraints \cite{davidson2007survey,wagstaff2001constrained}.}\label{fig:figure15}
\end{figure}

@@

The already mentioned constrained clustering applications are the most studied in the literature, although the reader may have the impression that constrained clustering is no longer used in real-world problems. To prevent that we provide a list of some more recent applications of constrained clustering to modern problems: advanced robotics navigation systems \cite{semnani2016constrained}, applied marketing \cite{seret2014new}, obstructive sleep apnea analysis \cite{mai2018evolutionary}, handwritten digits classification \cite{li2015scalable}, Internet traffic classification \cite{wang2014internet}, electoral district designing, \cite{brieden2017constrained}, semantically similar relations identification \cite{wang2015constrained}, cooperative team building \cite{yang2014team}, video face clustering \cite{zhou2014video}, image clustering \cite{habashi2017semi} and storage location assignment in warehouses \cite{yang2016constrained} among others.

@@

	\item Increased accuracy in label predictions by generating constraints based on a subset of labeled instances.

	\item Obtaining clusters with adaptable geometry to each problem.

@@

Given $X = \{x_1 \cdots x_n\}$, a large set of unlabeled instances, and $L = \{(x_{n+1}, l_{n+1}) \cdots (x_{n+|L|}, l_{n+|L|})\}$, a small set of labeled instances, it is common to choose two elements from $L$ (with replacement) and set a \acs{ML} constraint between them if they belong to the same class, or a \acs{CL} constraints otherwise. An appropriate method to evaluate the results provided by a clustering method is to measure the level of accuracy it achieves when predicting dataset $X$ labels. This normally requires specifying the number of desired clusters equal to the number of known classes in $X$ ($k = k^*$). Methods such as the RandIndex \cite{rand1971objective} or its variants, which will be discussed later, are used to measure the degree of agreement between two given partitions (clusterings).

In \cite{wagstaff2000clustering}, where constraints are generated as described above, it is shown that accuracy is increased up to a 20\% over classic methods when averaging results over different constraint sets.

@@

This rule, however, is not true in all cases, as in datasets such as \textit{Tic-Tac-Toe Endgame}, no increase in predictions is achieved regardless of the number of constraints used. A possible explanation for these exceptions is based on the fact that setting $k = k^*$ may not be appropriate in these cases.

@@

Although, as we have seen, incorporating constraints to clustering methods brings benefits in some applications, there are two main drawbacks that are discussed below, as well as possible solutions to them.

@@

Introducing constraints in clustering changes the problem it solves, which becomes: \textit{Find the best partition that satisfies all constraints}. This way, if constraints are not properly specified or if the extraction methods are inadequate, we can find that the constraints contradict each other, which means that there is no partition that satisfies them all. For instance, there is no partition that satisfies the constraints $C_=(x_i,x_j)$ and $C_{\neq}(x_i,x_j)$, regardless of the value of $k$. The same happens for $k = 2$, and the constraints $C_{\neq}(x_i, x_j)$, $C_{\neq}(x_j, x_q)$ and $C_{\neq}(x_i, x_q)$. Definition \ref{def:FeasibilityProblem} formalizes the feasibility problem for non-hierarchical instance-level constrained clustering:

@@

As shown in Table \ref{tab:CCComplexity}, the use of \acs{CL} constraints increases the constrained clustering complexity level to $\mathbf{NP}$-complete and therefore constrained clustering is intractable. Intuitively it can be easily understood that if finding a single partition that satisfies the constraints is a complex problem, finding the best of them is harder.

@@

	\textbf{Knowing a feasible solution exists does not help us find it.} The consequences of this result on the constrained clustering complexity imply that, even if there is a feasible partition, it will not be easy to find, speaking in terms of algorithmic complexity \cite{davidson2007survey}.
	\label{ob:FeasibleSolution}

@@

In \cite{wagstaff2002intelligent} and \cite{davidson2007hierarchical} is shown that, even in the number of clusters in the output partition is set equal to the number of classes in the dataset ($k = k*$), which guarantees that a feasible solution does exists, simple algorithms like COP-kmeans \cite{wagstaff2001constrained} may not converge due to the feasibility problem.

@@

In constrained clustering it is assumed that constraints are a guide for an algorithm to find de desired data partition. So, it is reasonable to think that the more additional information (constraints) we have available, the closer the result we get to the one we are looking for, as Observation \ref{ob:AccuracyIncrease} stated. However, and despite the stipulations of this observation, we find cases in which, even generating the constraints without noise and on the basis of the true labels, there are constraint sets that, far from improving the results, worsen them considerably \cite{davidson2006proceedings}. This seems to disagree with the Observation \ref{ob:AccuracyIncrease}, however, let us remember that it refers to the average case, and not to particular cases.

@@

This result allows you to transfer many of the properties of the graph coloring problem to the constrained clustering problem. For example, Brook's theorem states that graph coloring is simple when the number of available colors ($k$ in our case) is greater than the maximum degree of the graph.

>>>>

Once the basics concepts about clustering have been introduced, we can focus on defining the concepts specific to constrained clustering. We give example applications and highlight its advantages and disadvantages. To this end we take as our main reference the survey presented in \cite{davidson2007survey}. We describe the motivation for constrained clustering in Section \ref{sec:CCMotiv}. We formalize the concept of constraints and discuss their use in Sections \ref{sec:ConstDef} and \ref{sec:ConstUse} respectively. Afterwards a list of applications of constrained clustering is presented in Section \ref{sec:CCApplications}. Benefits and problems, as well as solutions for these problems, are discussed in Sections \ref{sec:ConstraintsBenefits} and \ref{sec:ConstraintsProblems} respectively.

@@

As we discussed in Chapter \ref{ch:IntroClustering}, unsupervised clustering methods are useful for structuring data referring to a particular area. An example of this can be found in text classification; In \cite{cohn2003semi} a problem proposed by Yahoo! is addressed. The problem consists in, given a large number of text documents, grouping them according to a taxonomy in which documents with similar topics are conceptually nearby. To solve this problem, unsupervised clustering methods are useful, as there is limited initially available information on the problem. However, in \cite{wagstaff2001constrained} it is shown that applying unsupervised clustering to certain problems, such as grouping \acf{GPS} data in such a way that clusters define the lanes of a road, does not produce significant results, as the clusters obtained are far from the elongated shape that would be expected. To tackle the problem, they introduced a new element in clustering, the \textit{constraint}, which made it possible to include domain knowledge to guide clustering methods towards the expected results. It was enough to indicate that the lanes of the road on which the vehicles circulate are four meters wide, and therefore any vehicle that is at a distance of more than 4 meters from another, perpendicular to the motion direction, must be located in a different cluster.

We are now in a new scenario: it is possible to incorporate additional information into the clustering process, in addition to that contained in the dataset itself, to guide it in the formation of the output partition and obtain more accurate results. This new approach is known as constrained clustering, an \acf{SSL} paradigm, unlike traditional clustering methods which fall into the category of unsupervised learning.

@@

The new type of information that we incorporate into clustering is given in the form of instance-level constraints. Their purpose is to specify whether two instances ($x_i$ and $x_j$) from the dataset ($X$) must be in the same cluster or, on the contrary, they must be in separate clusters.

@@

Although they may seem simple, constraints defined as above have some very interesting properties. \acs{ML} constraints are an example of an equivalence relation, and therefore are symmetrical, reflexive and transitive. Observation \ref{ob:MLConstraintsRelation} formalizes this concept. On the other hand, and although \acs{CL} do not constitute an equivalence relation, they feature other properties, formalized in Observation \ref{ob:CLConstraintsRelation}.

@@

A clear example of the use of constraints can be found in clustering applications where there are distance measurement limitations, as in the case of \acs{GPS} data. Thus, if we want instances forming two clusters to be separated by a distance greater or equal to $\delta$, it is enough to set \acs{ML} constraints between all instances separated by less than $\delta$ units. Similarly, if we want the diameter of the clusters to be at most $\epsilon$, we must set \acs{CL} constraint between instances separated by more than $\epsilon$ units. Figure \ref{fig:DistanceConstraints} shows a graphical representation of these two types of constraints.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.45]{gfx/ConstClust/RestriccionesDeltaEpsilon.png} 
	\caption[\textit{delta} $(\delta)$ and \textit{epsilon} $(\epsilon)$ constraints.]{\textit{delta} $(\delta)$ and \textit{epsilon} $(\epsilon)$ constraints \cite{davidson2007survey}.}\label{fig:DistanceConstraints}
\end{figure}

@@

While supervised learning involves knowing the label associated with each instance, \acs{SSL} has only a subset of labeled instances. On the other hand, in many domains the available information refers to relations between instances, and not to the specific class to which they belong. Moreover, in interactive clustering systems, a user who is not expert in the problem domain will probably be able to provide information in the form of constraints such as \acs{ML} and \acs{CL}, instead of providing information on what particular class certain instances belong to \cite{cohn2003semi,davidson2007hierarchical}.

Typically, constraints are incorporated into clustering problems in two ways. They can be used to alter the rules of assignment of instances to clusters of the method in question, so that the solution satisfies as many constraints as possible. Alternatively, it is possible to train the distance function used by the method based on the constraints, either before or during the application of the method. In any case, the initialization phase can take the constraints into account, so that instances appearing together in \acs{ML} constraints will be placed in the same cluster, and those between which there is a \acs{CL} constraint will be placed in different clusters. Based on this distinction, we identify two ways of approaching the problem: those based on explicit constraints (\textit{constraints-based}) and those based on distances (\textit{distance-based}).

@@

	\item Initializing clusters based on constraints inferred from a set of labeled instances \cite{basu2002semi}.

@@

	\caption[Constraints and clustering over a dataset.]{Constraints and clustering over a dataset \cite{davidson2007survey}.} \label{fig:ConstClustOverDataset}
\end{figure}

@@

In distance-based approaches, classic clustering methods that use a distance measurement are modified so that the distance measurement incorporates the constraints. In this context, constraint satisfaction means that instances co-occurring in \acs{ML} constraints are placed together in the space, and those co-occurring in \acs{CL} are separated.

Figure \ref{fig:ConstrainsAndMetricLearned} shows a possible clustering based on a metric learned from the constraints. It should be noted that in Figure \ref{fig:MetricLearned} the data space has been compressed on the vertical axis and expanded on the horizontal axis to match the learned distance metric.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Constraints over a dataset.]
	{\includegraphics[width=.45\linewidth]{gfx/ConstClust/InputInstancesAndConst2}} 
	\hspace{0.5cm}
	\subfloat[Clustering based on a metric learned from constraints.]
	{\includegraphics[width=.45\linewidth]{gfx/ConstClust/MetricaAprendida}
	\label{fig:MetricLearned}}
	\caption[Clustering based on a metric learned from constraints.]{Clustering based on a metric learned from constraints \cite{davidson2007survey}.} \label{fig:ConstrainsAndMetricLearned}
\end{figure}

@@

This section shows some application cases in which constrained clustering has turned out to be a more useful tool than fully unsupervised clustering. For each case we will analyze how the constraints were obtained and how they improve the resulting clustering.

@@

Figure \ref{fig:CMUFacesDatabase} shows a sample from the \acsfont{CMU} (Carnegie Mellon University) faces dataset, where the task is to group faces according to different criteria (in this case, face orientation).

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Profile faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras1}} \quad
	\subfloat[Front faces.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras3}} \quad
	\subfloat[Faces up.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Caras2}} \quad
	\caption[\acsfont{CMU} faces database.]{\acsfont{CMU} faces database \cite{davidson2007survey}.}\label{fig:CMUFacesDatabase}
\end{figure}

The method used to obtain the constraints is one of the most popular in the literature: let the number of clusters of the resulting partition be equal to the number of classes in the database, and generate the constraints from a subset of labeled instances; in other words, if two instances have different labels, set a \acs{CL} constraint between them, and otherwise set an \acs{ML} one. Thus, between the images shown in Figure \ref{fig:FacesDatabaseCL} \acs{CL} constraints are set, since, although they depict the same person, they do not have the same orientation.

@@

Figure \ref{fig:AiboRobotClustSys} shows another set of image data to which constrained clustering techniques are applied. In this case, the task is to perform object recognition to incorporate the method into the navigation system of the Aibo robot \cite{davidson2005clustering}. Distance constraints such as $\delta$ and $\epsilon$ are used as described in Figure \ref{fig:DistanceConstraints}. In this way, well differentiated clusters are obtained and therefore they are useful for the path-finding tasks performed by the robot during navigation.

\begin{figure}[bth]
	\myfloatalign
	\subfloat[Original image.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo1}} \quad
	\subfloat[Classic clustering.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo2}} \quad
	\subfloat[Constrained clustering.]
	{\includegraphics[width=.3\linewidth]{gfx/ConstClust/AnalisisImagenes/Aibo3}} \quad
	\caption[Clustering method used in Aibo robot navigation system.]{Clustering method used in Aibo robot navigation system \cite{davidson2007survey,davidson2005clustering}.}\label{fig:AiboRobotClustSys}
\end{figure}

@@

Video databases are one example where constraints can be generated directly from the data domain, especially if space-time metadata is available \cite{yan2006discriminative}. In time-sequenced data it is possible to set \acs{ML} constraints between groups of pixels of frames close in time. This is especially useful when the task is to perform object recognition based on clustering and segmentation. It is also possible to add \acs{CL} constraints to image segments in the same snapshot, as the probability of them being associated with the same object after segmentation is low. In fact, in video analysis problems there are a variety of constraint extraction methods \cite{yan2006discriminative}. Figure \ref{fig:ConstExtractionVideo} shows some examples.

@@

In Figure \ref{fig:ConstExtractionVideo}, image \ref{fig:ConstExtractionVideoA} corresponds to constraints extracted from tracking a person over a period of time; \ref{fig:ConstExtractionVideoB} corresponds to spatial constraints associating two objects located in the same frame; image \ref{fig:ConstExtractionVideoC} corresponds to constraints obtained through facial recognition, and \ref{fig:ConstExtractionVideoD} to those provided by the user.

With so many constraint extraction methods, one might ask: What happens if too many constraints are imposed? Does this make the problem over-constrained? In Section \ref{sec:ConstraintsProblems} we will address these questions.

@@

In gene clustering based on microarray data, genes are represented by their expression profile in different experiments and grouped using different constrained clustering methods. Figure \ref{fig:GeneticApp} shows an example: these are \acs{ML} constraints set between genes based on co-occurrence data stored in the protein interaction database, which contains information about which genes---and their associated proteins---are present in the same cell processes \cite{xenarios2001dip}. This information can be used to improve results provided by clustering methods. \cite{segal2003discovering}.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/Genetica/Genes} 
	\caption[Gene clustering based on microarray data.]{Gene clustering based on microarray data \cite{davidson2007survey}.}\label{fig:GeneticApp}
\end{figure}

@@

In content classification tasks, the goal is to automatically split large amounts of documents into groups or clusters. In this case it is possible to extract constraints from multiple auxiliary resources. For example, if two documents are in the same directory, an \acs{ML} constraint could be set between them. This way, it is possible to adjust the resulting clustering to suit a particular criterion, such as creating a hierarchy of documents akin to the way they are organized in the original directory structure.

@@

Constrained clustering is very useful in web search data processing. Here, the aim is to automatically group the results of an ambiguous query to the search engine into clusters of \acsp{URL}, which represent the queried concept in different contexts. In this area it is possible to extract the constraints from previous searches performed by other users, so that an \acs{ML} constraint is set between \acsp{URL} visited in the same user session. Clustering using these constraints can help bias search results towards user preferences.

@@

In certain audio analysis tasks, the number of object classes present in the data may not be known, although constraints can be extracted directly from the data domain. This happens, for example, when applying clustering to speaker recognition in a conversation \cite{bar2003learning}. In this case the number of participants is not known a priori, but it is easy to detect if two speakers are different or similar and to set constraints accordingly.

\subsection[\acsfont{GPS} Data]{GPS Data} \label{sec:GPSApp}

As mentioned at the beginning of Chapter \ref{ch:ConstrainedClustering}, constrained clustering can be applied to \acs{GPS} data in order to identify the lane each vehicle is in, as shown in Figure \ref{fig:GPSClustData}. Each instance is represented by the position it occupies on the road in two-dimensional Cartesian coordinates $(x,y)$ obtained from \acs{GPS} data. Figure \ref{fig:figure15} graphically shows this representation of the data. Note that multiple instances can refer to the same vehicle at different times.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/ConstClust/GPS/Coches} 
	\caption[\acs{GPS} data use.]{\acs{GPS} data use \cite{davidson2007survey, wagstaff2001constrained}.}\label{fig:GPSClustData}
\end{figure}

In this domain, real clusters have an elongated shape on the horizontal axis and are aligned perpendicularly to the motion direction. To produce clusters with this shape, we can make use of constraints. We will set \acs{CL} constraints between those instances more than four meters apart perpendicularly to the motion direction---since the lanes have a maximum width of four meters---and \acs{ML} constraints between those instances that present continuity in the motion direction axis, since it is likely that the vehicles they represent are in the same lane. This clustering model has proven to be very useful in real-time navigation \cite{wagstaff2001constrained}, allowing for the user to be notified when to change lanes, or when not to.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.32]{gfx/ConstClust/GPS/Instancias} 
	\caption[Clustering of \acs{GPS} data with no constraints.]{Clustering of \acs{GPS} data with no constraints \cite{davidson2007survey,wagstaff2001constrained}.}\label{fig:figure15}
\end{figure}

@@

The aforementioned applications are the most studied in the literature, although the reader may have the impression that constrained clustering is no longer used in real-world problems. To prevent that, we provide a list of more recent applications of constrained clustering to modern problems: advanced robotics navigation systems \cite{semnani2016constrained}, applied marketing \cite{seret2014new}, obstructive sleep apnea analysis \cite{mai2018evolutionary}, handwritten digits classification \cite{li2015scalable}, Internet traffic classification \cite{wang2014internet}, electoral district designing, \cite{brieden2017constrained}, semantically similar relations identification \cite{wang2015constrained}, cooperative team building \cite{yang2014team}, video face clustering \cite{zhou2014video}, image clustering \cite{habashi2017semi} and storage location assignment in warehouses \cite{yang2016constrained}, among others.

@@

	\item Increased accuracy in label prediction by generating constraints based on a subset of labeled instances.

	\item Obtention of clusters with adaptable geometry to each problem.

@@

Given $X = \{x_1 \cdots x_n\}$, a large set of unlabeled instances, and $L = \{(x_{n+1}, l_{n+1}) \cdots (x_{n+|L|}, l_{n+|L|})\}$, a small set of labeled instances, it is common to choose two elements from $L$ (with replacement) and set an \acs{ML} constraint between them if they belong to the same class, or a \acs{CL} constraint otherwise. An appropriate way to evaluate the results provided by a clustering method is to measure the level of accuracy it achieves when predicting dataset $X$ labels. This normally requires the number of desired clusters to be equal to the number of known classes in $X$ ($k = k^*$). Methods such as the RandIndex \cite{rand1971objective} or its variants, which will be discussed later, are used to measure the degree of agreement between two given partitions (clusterings).

In \cite{wagstaff2000clustering}, where constraints are generated as described above, it is shown that accuracy is increased up to 20\% over classic methods when averaging results over different constraint sets.

@@

This rule, however, is not always true, as in datasets such as \textit{Tic-Tac-Toe Endgame}, no increase in correct predictions is achieved regardless of the number of constraints used. A possible explanation for these exceptions is that setting $k = k^*$ may not be appropriate in these cases.

@@

Although, as we have seen, incorporating constraints into clustering methods brings benefits in some applications, there are two main drawbacks that are discussed below, as well as possible solutions to them.

@@

The introduction of constraints into clustering changes the problem it solves, which becomes: \textit{Find the best partition that satisfies all constraints}. This way, if constraints are not properly specified or if the extraction methods are inadequate, we can end up with contradictory constraints, which means that there is no partition that satisfies them all. For instance, there is no partition that satisfies the constraints $C_=(x_i,x_j)$ and $C_{\neq}(x_i,x_j)$, regardless of the value of $k$. The same happens for $k = 2$, and the constraints $C_{\neq}(x_i, x_j)$, $C_{\neq}(x_j, x_q)$ and $C_{\neq}(x_i, x_q)$. Definition \ref{def:FeasibilityProblem} formalizes the feasibility problem for non-hierarchical instance-level constrained clustering:

@@

As shown in Table \ref{tab:CCComplexity}, the use of \acs{CL} constraints increases the complexity level of constrained clustering to $\mathbf{NP}$-complete and therefore constrained clustering is intractable. Intuitively, it can be easily understood that if finding a single partition that satisfies the constraints is a complex problem, finding the best of them is even harder.

@@

	\textbf{Knowing a feasible solution exists does not help us find it.} The consequences of this result for the complexity of constrained clustering imply that, even if there is a feasible partition, it will not be easy to find in terms of algorithmic complexity \cite{davidson2007survey}.
	\label{ob:FeasibleSolution}

@@

In \cite{wagstaff2002intelligent} and \cite{davidson2007hierarchical} is shown that, even if the number of clusters in the output partition is equal to the number of classes in the dataset ($k = k^*$), which guarantees that a feasible solution does exist, simple algorithms like \acf{COPKM} \cite{wagstaff2001constrained} may not converge due to the feasibility problem.

@@

In constrained clustering it is assumed that constraints are a guide for an algorithm to find the desired data partition. So, it would be reasonable to think that the more additional information (constraints) we have available, the closer the result will be to the one we are looking for, as Observation \ref{ob:AccuracyIncrease} stated. However, and despite the stipulations of this observation, we find cases where, even generating the constraints without noise and according to the true labels, there are constraint sets that, far from improving the results, worsen them considerably \cite{davidson2006proceedings}. This seems to disagree with Observation \ref{ob:AccuracyIncrease}; however, let us remember that it refers to the average case, and not to particular cases.

@@

This result allows us to transfer many of the properties of the graph coloring problem to the constrained clustering problem. For example, Brook's theorem states that graph coloring is simple when the number of available colors ($k$ in our case) is greater than the maximum degree of the graph.
